---
title: "Process for understanding data"
format: html
code-fold: false
---

The following page details our process for wrangling the data to create the graphs present in our report.

```{r}
#| label: Initialize project
#| warning: false
library(tidyverse)
library(here)
library(rnaturalearth)
library(sf)
library(RColorBrewer)
library(viridisLite)
```


Importing the original data set
```{r}
Invasive_Species <- read_csv(here("data", "Invasive_Species.csv"))
# Data source: Vermont Center for Geographic Information (VCGI), https://catalog.data.gov/dataset/invasive-species-812ab
```

Creating a list of the invasive species that are most represented in the original data set
```{r}
TopInvasive <- Invasive_Species |>
  group_by(InvasiveName) |>
  count() |>
  filter(n > 6)
```

Joining that list with the main data set and adding a "year" column
```{r}
Top_Invasive_Species <- Invasive_Species |>
  inner_join(TopInvasive, by = join_by(InvasiveName)) |>
  mutate(year = str_sub(ObservationDate, -4, str_length(ObservationDate)))
```

Importing data set shape map
```{r}
invasive_boundaries <- read_sf(here("data/Invasive_Species_Shapefiles", "Invasive_Species.dbf"))
```

Joining shape map with Top_Invasive_Species data set and filtering for Hemlock Woolly Adelgid (HWA)
```{r}
invasive_boundaries |>
  mutate(GrossArea = as.numeric(GrossArea)) |>
  inner_join(Top_Invasive_Species) |>
  filter(InvasiveNa == "Hemlock Woolly Adelgid")
```

